---
title: "Replicating Freedman (2012): does NMTC money jump at the eligibility cutoff?"
author: "QSS Lab — Portugal pod"
date: "Summer 2026"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Read this first

This notebook walks you through replicating one result from one published paper:

> Freedman, Matthew (2012), "Teaching new markets old tricks: The effects of
> subsidized investment on low-income neighborhoods," *Journal of Public
> Economics* 96(11–12), 1000–1014.

The one result is the paper's **first stage**: census tracts whose median family
income sits just below 80% of their area's median receive discontinuously more New
Markets Tax Credit investment than tracts just above. We are not redoing the whole
paper. We are rebuilding its engine and checking that it turns over.

The structure follows the worksheet: data, sample, specification, run, compare,
reconcile. A few rules that will save you time:

1. Work top to bottom. Part 2 is a dress rehearsal on fake data and everything
   after it reuses the pattern you build there. Don't skip it.
2. Run every chunk yourself. Reading code and running code are different skills
   and only one of them is the one you're here to learn.
3. Stuck more than 30 minutes, ask. That's the lab rule. The people who fail at
   this are the ones who sit with a bug in silence for a week.
4. Boxes marked **CHECKPOINT** say what you should be seeing. When your numbers
   disagree, stop and find out why. The discrepancies are the curriculum.

# Setup

Install [R](https://cran.r-project.org/), then [RStudio Desktop](https://posit.co/download/rstudio-desktop/),
then open this file in RStudio. The chunk below installs what's missing and loads it.

```{r packages}
pkgs <- c(
  "tidyverse",  # data manipulation and plotting
  "scales",     # readable axis labels
  "broom",      # turns regression output into data frames
  "estimatr"    # lm_robust: regression with heteroskedasticity-robust SEs
)
to_install <- setdiff(pkgs, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)

library(tidyverse)
library(scales)
library(broom)
library(estimatr)
```

## The papers

Both PDFs are already in `papers/`, next to this file:

- `freedman_2012_teaching_new_markets_old_tricks_wp.pdf` — the paper you're replicating
- `baum_snow_marion_2009_lihtc_wp.pdf` — the backup paper, for later

These are the authors' ungated working-paper versions. If you have library access,
also pull the published *JPubE* version of Freedman, since tables sometimes move
between working paper and journal, and the published table is the one your Part 4
comparison should target.

Before Part 3, read Freedman: sections 1–2 (the program and the eligibility rule)
and section 3 (data) carefully, the first-stage table and figure very carefully —
finding *which* table and figure hold the first stage is your first worksheet entry —
and the rest on a skim. While reading, fill in the worksheet's paper card: question,
identification idea, data, sample rules, specification, headline numbers.

# The idea, in plain words

The New Markets Tax Credit (enacted 2000) gives investors a 39% federal tax credit
for equity put into community development entities, which must invest it in
businesses and real estate in low-income communities. Billions of dollars flow
through it.

A census tract is a "low-income community" if its poverty rate is at least 20%,
*or* its median family income is at most 80% of the area median — state median for
non-metro tracts, the larger of state and metro median for metro tracts. Cross either
threshold and you're eligible for that capital.

A tract at 79.9% of area income and a tract at 80.1% are, for any economic purpose,
the same place. But one is eligible for subsidized capital and the other isn't.
Comparing tracts just below the line to tracts just above is about as close to a
randomized experiment as observational data gets — that's a regression discontinuity
design.

Of course, none of it matters unless the rule actually moves money. Do just-below
tracts really get more NMTC investment than tracts just above? That jump is what
we're estimating — and it's called the first stage.

One more thing: "does public money pull private money into places it otherwise
skips?" is the question Katia's COMPETE2020 project asks about Portugal. You're
building the US benchmark for the thing you'll help her measure.

Vocabulary you'll need: *running variable* (the income ratio the rule is applied
to), *cutoff* (0.80), *window* or *bandwidth* (how close to the cutoff a tract must
be to count as "close"), *first stage* (effect of crossing the cutoff on treatment),
*QLICI* (one NMTC loan or investment — one row in the transaction data).

# Part 2 — Dress rehearsal on fake data

Before touching real data, build a small fake world where you know the truth and
check that your code recovers it. Keep this habit forever: code that can't find an
effect you planted on purpose won't find one in the wild either.

## Simulate a world with a known jump

```{r simulate}
set.seed(45)   # for §45D, the NMTC statute. Reproducibility is not optional.

n_tracts <- 4000

fake <- tibble(
  geoid     = sprintf("%011d", sample(1e9, n_tracts)),
  mfi_ratio = runif(n_tracts, min = 0.40, max = 1.20)
) |>
  mutate(
    below_cutoff = mfi_ratio <= 0.80,
    # the truth we plant: crossing the cutoff raises the chance of getting a
    # deal by 8 points; richer tracts get a bit less in general
    prob_invest  = 0.05 + 0.08 * below_cutoff + 0.10 * (1 - mfi_ratio),
    got_nmtc     = rbinom(n_tracts, size = 1, prob = prob_invest),
    qlici_amount = if_else(got_nmtc == 1, round(exp(rnorm(n_tracts, 15, 1))), 0)
  )

glimpse(fake)
```

> **CHECKPOINT.** 4,000 rows, 6 columns. `qlici_amount` is zero for most tracts and
> in the millions for some. (The real first stage you'll estimate in Part 3 is a
> good deal smaller than the 8 points we planted here; the fake world is tuned to
> be easy to see, not to be realistic.)

## Look before you regress

The most convincing object in any RD paper is the picture: bin the running
variable, average the outcome within bins, look for the step.

```{r rd-plot-fake, fig.width=8, fig.height=5}
fake |>
  mutate(bin = cut_width(mfi_ratio, width = 0.02, center = 0.01)) |>
  group_by(bin) |>
  summarise(bin_mid = mean(mfi_ratio), share = mean(got_nmtc), n = n()) |>
  ggplot(aes(bin_mid, share)) +
  geom_point(aes(size = n), alpha = 0.7) +
  geom_vline(xintercept = 0.80, linetype = "dashed", color = "red") +
  scale_y_continuous(labels = percent) +
  labs(
    title = "Fake world: share of tracts with any NMTC deal",
    x = "Tract MFI / area MFI", y = "Share with any deal", size = "Tracts in bin"
  ) +
  theme_minimal()
```

> **CHECKPOINT.** A gentle downward drift with a visible step down exactly at the
> dashed line. That step is your planted 8 points.

## Measure the jump

Keep tracts near the cutoff and let the outcome depend on a below-cutoff dummy plus
a straight line in the running variable that may differ on each side. The dummy's
coefficient is the jump.

```{r rd-reg-fake}
window <- fake |>
  filter(mfi_ratio >= 0.70, mfi_ratio <= 0.90) |>
  mutate(dist = mfi_ratio - 0.80)

first_stage_fake <- lm_robust(
  got_nmtc ~ below_cutoff + dist + below_cutoff:dist,
  data = window
)

tidy(first_stage_fake) |> select(term, estimate, std.error, p.value)
```

> **CHECKPOINT.** `below_cutoffTRUE` lands near 0.08, within about two standard
> errors. If it does, your pipeline works and you've run your first regression
> discontinuity.

## Two experiments before you move on

Each one teaches an argument you'll need later. Write a sentence about what you see.

1. Shrink the window to 0.75–0.85 and re-run. The estimate should stay near 0.08
   while the standard error grows. Fewer tracts, same truth — that trade-off is the
   bandwidth debate in every RD paper.
2. Placebo: pretend the cutoff is 0.90 (re-center `dist`, window 0.85–0.95). The
   "jump" should be near zero. Why does that reassure you about the real one?

```{r experiments, eval=FALSE}
# your code here — copy the chunk above and modify it
```

# Part 3 — The real thing

The `data/` folder next to this notebook ships with three committed files, built
from public sources in June 2026. Read `data/DATA_DICTIONARY.md` before loading
anything — two minutes there saves an hour here.

One paragraph of honesty about where this data comes from. The CDFI Fund re-reports
all historical deals on whatever census geography is current, so the file the lab
holds (FY2022 release) puts every deal on **2020** tracts — useless against a year-2000
income rule — and the oldest surviving tract-level release (FY2017) uses **2010**
tracts. The deals here are therefore mapped from 2010 back to 2000 boundaries with
the Census population crosswalk, and the running variable is rebuilt from Census 2000
sources the way Freedman describes. The dictionary documents every choice. Freedman
worked from the confidential file on native 2000 tracts, so expect your estimate to
be attenuated relative to his — and expect to explain *why* in Part 4. This is what
the worksheet means by data vintage, and you'll meet it again in every project.

## The transaction file: cleaning practice

```{r load-tx}
tx <- read_csv("data/nmtc_transactions_fy2003_2017.csv",
               col_types = cols(tract_geoid_2010 = col_character()))
glimpse(tx)
```

A tract GEOID is an 11-digit label, 2 state + 3 county + 6 tract. Read as a number,
`01001020100` silently becomes `1001020100` and never matches anything again. IDs
are text. That's why the `col_types` argument is there; delete it once, read the
file, and look at what happens to Alabama.

Now interrogate the file and write down what you find — this is the worksheet's
"data" section:

```{r clean-tx}
nrow(tx)                          # 13,880
count(tx, year) |> print(n = 20)  # look closely at the early years
summary(tx$qlici_amount)
mean(tx$multi_tract == "YES")     # what does a multi-tract deal even mean here?
```

> **CHECKPOINT.** 13,880 rows. The year counts run 2003–2017 — except for five rows
> dated 2001 and 2002, before the program made its first allocations. Real
> administrative data does this. Decide what to do with them and write the decision
> down; there is no answer key, only defensible choices.

Freedman's sample covers the program's early years. Check section 3 of the paper
for his exact window, then build yours:

```{r filter-years}
paper_years <- 2003:2007   # verify this against the paper before trusting it
tx_sample <- tx |> filter(year %in% paper_years)
nrow(tx_sample)            # 2,666
```

## The running variable

```{r load-elig}
elig <- read_csv("data/eligibility_2000.csv",
                 col_types = cols(tract_geoid = col_character(),
                                  metro_id = col_character()))

nrow(elig)                                            # 65,443 tracts
mean(elig$mfi_ratio <= 0.80, na.rm = TRUE)            # ~0.37 income-eligible
mean(elig$mfi_ratio <= 0.80 | elig$poverty_rate >= 0.20, na.rm = TRUE)  # ~0.39
mean(elig$metro)                                      # ~0.75
```

> **CHECKPOINT.** 65,443 tracts; 39.0% eligible by income or poverty. The CDFI
> Fund's own published figure for this era is "about 39 percent of tracts." When a
> number you built from scratch hits an external benchmark, say so in your write-up;
> that's what credibility looks like.

Watch out for a trap that quietly ruins this replication: eligibility is income ≤ 80%
**or** poverty ≥ 20%. A tract at 85% of area income with 22% poverty is eligible.
Leave those tracts in and your "ineligible" side is contaminated and the jump
shrinks. Find how Freedman handles this in the paper, then build the clean sample:

```{r or-clause}
elig_clean <- elig |>
  filter(poverty_rate < 0.20,          # income is now the only eligibility margin
         !startsWith(tract_geoid, "72"))  # Freedman's sample is the 50 states + DC
nrow(elig_clean)
```

## The merge

The transaction data only contains tracts that got deals. The analysis file needs
every tract, with explicit zeros for the untreated — the universe is the left table.
Deals are aggregated by tract for the sample years in
`data/nmtc_tract_qlici_2003_2007.csv`, which also carries each deal-tract's mapping
to 2000 boundaries. (Exercise for later: rebuild that file from `tx_sample`
yourself and check you get 1,255 tracts.)

```{r merge}
deals <- read_csv("data/nmtc_tract_qlici_2003_2007.csv",
                  col_types = cols(tract_geoid_2010 = col_character(),
                                   tract_geoid_2000 = col_character())) |>
  group_by(tract_geoid = tract_geoid_2000) |>
  summarise(n_deals = sum(n_deals), total_qlici = sum(total_qlici))

analysis <- elig_clean |>
  left_join(deals, by = "tract_geoid") |>
  mutate(
    n_deals     = replace_na(n_deals, 0),
    total_qlici = replace_na(total_qlici, 0),
    got_nmtc    = as.integer(n_deals > 0)
  )

mean(analysis$got_nmtc)   # small — most tracts never see a deal
```

Some deal tracts won't appear in `analysis`. Account for them: poverty-route tracts
you dropped on purpose, Puerto Rico, stray GEOIDs. "I can explain every row
I dropped" is the difference between a replication and a vibe.

## Run the first stage

Identical shape to the rehearsal — that was the point of the rehearsal.

```{r first-stage}
window_real <- analysis |>
  filter(mfi_ratio >= 0.70, mfi_ratio <= 0.90) |>
  mutate(below_cutoff = mfi_ratio <= 0.80,
         dist = mfi_ratio - 0.80)

nrow(window_real)
window_real |> group_by(below_cutoff) |> summarise(n = n(), share_any = mean(got_nmtc))

fs_any <- lm_robust(got_nmtc ~ below_cutoff + dist + below_cutoff:dist,
                    data = window_real)
fs_amt <- lm_robust(total_qlici ~ below_cutoff + dist + below_cutoff:dist,
                    data = window_real)

tidy(fs_any) |> select(term, estimate, std.error, p.value)
tidy(fs_amt) |> select(term, estimate, std.error, p.value)
```

> **CHECKPOINT.** With this folder's data and the sample rules above you should get
> N = 15,582; share with any deal 2.64% below the cutoff vs 0.22% above; and a
> below-cutoff jump of about **+0.017** (se ≈ 0.004, t ≈ 4.5) in the probability of
> any deal. Tracts just under the line are roughly twelve times more likely to see
> NMTC money. If your numbers differ, the bug is in your sample rules, not the data
> — go back one chunk at a time.

```{r rd-plot-real, fig.width=8, fig.height=5}
window_real |>
  mutate(bin = cut_width(mfi_ratio, width = 0.02, center = 0.01)) |>
  group_by(bin) |>
  summarise(bin_mid = mean(mfi_ratio), share = mean(got_nmtc), n = n()) |>
  ggplot(aes(bin_mid, share)) +
  geom_point(aes(size = n), alpha = 0.7) +
  geom_vline(xintercept = 0.80, linetype = "dashed", color = "red") +
  scale_y_continuous(labels = percent) +
  labs(
    title = "The real first stage: share of tracts with any NMTC deal, FY2003-2007",
    x = "Tract MFI / area MFI (2000 census)", y = "Share with any deal",
    size = "Tracts"
  ) +
  theme_minimal()
```

> **CHECKPOINT.** The plot should look like a cliff, not a slope: bins below 0.80
> sit between 2% and 4%, bins above collapse to under half a percent. Compare it to
> the first-stage figure in the paper.

# Part 4 — Compare and reconcile

This is the deliverable. Fill in the published numbers from the first-stage table
you identified in Part 0, and yours from Part 3:

```{r comparison}
comparison <- tribble(
  ~quantity,                            ~freedman_2012, ~ours,    ~same_sign, ~comparable_size,
  "Jump in P(any NMTC deal) at 0.80",   NA_real_,       NA_real_, NA,         NA,
  "Jump in NMTC $ at 0.80",             NA_real_,       NA_real_, NA,         NA,
  "N tracts in window",                 NA_real_,       NA_real_, NA,         NA
)
comparison
```

Success is not the identical number. You're using the public, de-identified release
crosswalked across census vintages; he used the confidential file on native 2000
tracts. Success is the same sign, a magnitude you can defend, and a written list of
reasons for every remaining gap. Work the suspects in order, cheapest first:

1. Sample window — same years? same bandwidth?
2. Sample rules — the poverty OR-clause; metro and non-metro both in; missing-income
   tracts; Puerto Rico.
3. Data — public subset vs confidential file; the 2010-to-2000 crosswalk
   (the `poppct10` column tells you which deal tracts mapped cleanly and which
   didn't — try re-running with only clean mappers).
4. Specification — local linear with his bandwidth, controls, clustering. Match one
   ingredient at a time and watch what moves.
5. Units — dollars vs logs vs an indicator; percent vs proportion.

The regression is one line. Knowing why two careful people get different numbers
from "the same" data is the skill this whole exercise exists to build.

# Part 5 — Where this goes next

Only after Part 4 is written up on the worksheet:

1. **One outcome.** Take one neighborhood outcome around the cutoff (tract poverty
   or home values, 2005–2009 ACS) and run the same regression on it. Now you've
   touched the paper's reduced form.
2. **The rural split.** Estimate the first stage separately for metro and non-metro
   tracts (the `metro` column is in the eligibility file). With this folder's data
   the non-metro jump in P(any deal) comes out *larger* than the metro one — but
   that's shares, and Ian's rural-gap question is about dollars mobilized. Which
   measure matters, and what do you find in dollars? This feeds a live lab paper.
3. **Rebuild the running variable yourself** from the Census API with the
   `tidycensus` package (ask Ian for the lab's Census API key; the dictionary
   describes the construction). If your homemade ratio matches `mfi_ratio`, you've
   independently verified the most fragile part of this whole folder.
4. **The Portugal question.** One page: what is the COMPETE2020 analogue of this
   design? Running variable, cutoff, data Katia would need. That's rung 5 — your
   first research proposal, and it's the one we actually want.

Then we talk about Baum-Snow & Marion (2009) — same skills, new program — or you
keep pulling on (4). Your call.
