---
title: 'Replicating Freedman (2012): does NMTC money jump at the eligibility cutoff?'
author: "QSS Lab — Portugal pod (Gabi, advised by Ian & Liz)"
date: "Summer 2026"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Read this first

Welcome! This notebook is the spine of your first replication. It walks the
worksheet steps — **data → sample → specification → run → compare → reconcile** —
for *one* result from one published paper:

> **Freedman, Matthew (2012), "Teaching new markets old tricks: The effects of
> subsidized investment on low-income neighborhoods,"** *Journal of Public
> Economics* 96(11–12), 1000–1014.

**The one result we are replicating (worksheet step 0):** census tracts whose
median family income falls *just below* 80% of their area's median receive
discontinuously more New Markets Tax Credit (NMTC) investment than tracts *just
above* that line. This is the paper's **first stage** — the engine that makes
everything else in the paper credible. We are *not* replicating the full paper.

A few ground rules:

1. **Work top to bottom.** Part 2 uses *fake* data that you generate yourself, so
   you can run the entire pipeline in 30 minutes before touching anything messy.
   Every later step reuses the pattern you build there.
2. **Run every chunk yourself** (the green ▶ button in RStudio). Reading code is
   not the same as running it.
3. **Stuck > 30 minutes → ask.** That's the lab rule. There are no dumb questions,
   only silent stuck people.
4. Boxes marked **CHECKPOINT** tell you what you should be seeing. If you see
   something else, stop and figure out why (or ask) — don't push on.

---

# Part 0 — Setup

## Install R and RStudio (once)

If you haven't already: install [R](https://cran.r-project.org/) first, then
[RStudio Desktop](https://posit.co/download/rstudio-desktop/). Open *this file*
(`freedman_first_stage.Rmd`) in RStudio.

## Install the packages we need (once)

```{r packages}
pkgs <- c(
  "tidyverse",  # data manipulation (dplyr) + plotting (ggplot2) — the workhorse
  "readxl",     # read Excel files (the CDFI Fund ships .xlsx)
  "janitor",    # clean_names(): turns "Project GEOID " into "project_geoid"
  "scales",     # nice axis labels ($, %, commas)
  "broom",      # tidy(): turns regression output into a data frame
  "estimatr"    # lm_robust(): regression with honest standard errors
)
to_install <- setdiff(pkgs, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)

library(tidyverse)
library(readxl)
library(janitor)
library(scales)
library(broom)
library(estimatr)
```

## Get the papers

Run this once. It creates a `papers/` folder next to this notebook and downloads
what it can; the Freedman PDF sits behind a click-through page, so the chunk opens
it in your browser instead — download it and save it into `papers/` yourself.

```{r get-papers, eval=FALSE}
dir.create("papers", showWarnings = FALSE)

# Baum-Snow & Marion (2009) — the backup paper, direct downloads:
download.file(
  "https://economics.brown.edu/sites/default/files/papers/2007-5_paper.pdf",
  "papers/baum_snow_marion_2007_wp.pdf", mode = "wb"
)
download.file(
  "https://www.novoco.com/public-media/documents/baumsnow-marion_lihtc-dec08.pdf",
  "papers/baum_snow_marion_2008_nearfinal.pdf", mode = "wb"
)

# Freedman (2012) — main paper. Two ungated copies; save one as
# papers/freedman_2012_nmtc.pdf
browseURL("https://ecommons.cornell.edu/handle/1813/89085")            # Cornell working paper
browseURL("https://papers.ssrn.com/sol3/papers.cfm?abstract_id=2180227") # SSRN copy
```

## Read the paper (before Part 3)

You don't need the paper for Parts 1–2. Before Part 3, read Freedman (2012):

- **Sections 1–2** (the program and the eligibility rule) carefully,
- **Section 3** (data) carefully — this is the section you are about to re-create,
- the **first-stage table and figure** carefully (find them! writing down *which*
  table/figure holds the first stage is your first worksheet entry),
- everything else (outcomes, robustness) on a skim for now.

As you read, fill in the worksheet's paper card: research question, identification
idea, data sources, sample restrictions, the exact first-stage specification, and
the headline first-stage numbers. **Copy the published coefficients into
Part 4's comparison table below** — they are your target.

---

# Part 1 — The idea, in plain words

**The program.** The New Markets Tax Credit (created 2000) gives investors a 39%
federal tax credit for putting equity into community development entities (CDEs),
which must invest it in businesses and real-estate projects in *low-income
communities*. Billions of dollars flow through it.

**The rule.** A census tract is a "low-income community" (eligible) if **either**

- its **poverty rate is ≥ 20%**, **or**
- its **median family income (MFI) is ≤ 80%** of the area median (the state
  median for non-metro tracts; the larger of state/metro median for metro tracts).

**The trick.** A tract at 79.9% of area MFI and a tract at 80.1% are basically
identical neighborhoods — but one is eligible for subsidized investment and the
other is not. Comparing tracts just below vs. just above the line is as close to
a randomized experiment as observational data gets. That's a **regression
discontinuity (RD)** design.

**The first stage.** None of this matters unless the rule actually moves money:
do just-below tracts really get more NMTC investment than just-above tracts? That
jump-at-the-cutoff is what we replicate.

**Why the lab cares.** "Does public money pull private money into places that
wouldn't otherwise get it?" is *exactly* the question the Portugal pod asks of
COMPETE2020. You're learning the US benchmark for the thing you'll help Katia
build.

> **Vocabulary you'll meet:** *running variable* (the MFI ratio — the continuous
> variable the rule is applied to), *cutoff* (0.80), *bandwidth/window* (how far
> from the cutoff we let tracts be and still count as "close"), *first stage* (the
> effect of crossing the cutoff on treatment, here NMTC dollars), *QLICI* (a
> qualified low-income community investment — one NMTC deal; one row in the data).

---

# Part 2 — Dress rehearsal on fake data

Before fighting real, messy data, we build a tiny fake world where we *know* the
truth, and check that our code recovers it. This is a habit worth keeping forever:
**if your code can't find an effect you put there on purpose, it won't find one in
real data either.**

## 2.1 Simulate a world with a known jump

```{r simulate}
set.seed(45)   # makes "random" numbers reproducible (45 for IRC §45D, the NMTC statute)

n_tracts <- 4000

fake <- tibble(
  geoid     = sprintf("%011d", sample(1e9, n_tracts)),  # 11-character tract IDs
  mfi_ratio = runif(n_tracts, min = 0.40, max = 1.20)   # tract MFI / area MFI
) |>
  mutate(
    below_cutoff = mfi_ratio <= 0.80,
    # The TRUTH we plant: crossing below the cutoff raises the chance a tract
    # gets any NMTC deal by 8 percentage points. Richer tracts also get a bit
    # less investment in general (the smooth, boring part).
    prob_invest  = 0.05 + 0.08 * below_cutoff + 0.10 * (1 - mfi_ratio),
    got_nmtc     = rbinom(n_tracts, size = 1, prob = prob_invest),
    qlici_amount = if_else(got_nmtc == 1, round(exp(rnorm(n_tracts, 15, 1))), 0)
  )

glimpse(fake)
```

> **CHECKPOINT.** `glimpse()` should show 4,000 rows and 6 columns, with
> `qlici_amount` equal to 0 for most tracts and in the millions for some.

## 2.2 *Look* before you regress

The single most convincing object in any RD paper is a picture: average the
outcome in narrow bins of the running variable and look for the jump.

```{r rd-plot-fake, fig.width=8, fig.height=5}
binned <- fake |>
  mutate(bin = cut_width(mfi_ratio, width = 0.02, center = 0.01)) |>
  group_by(bin) |>
  summarise(
    bin_mid    = mean(mfi_ratio),
    share_nmtc = mean(got_nmtc),
    n          = n()
  )

ggplot(binned, aes(x = bin_mid, y = share_nmtc)) +
  geom_point(aes(size = n), alpha = 0.7) +
  geom_vline(xintercept = 0.80, linetype = "dashed", color = "red") +
  scale_y_continuous(labels = percent) +
  labs(
    title    = "Fake world: share of tracts with any NMTC investment",
    subtitle = "Each dot = tracts in a 2-point bin of the income ratio. See the jump at 0.80?",
    x        = "Tract MFI / area MFI (running variable)",
    y        = "Share with any NMTC deal",
    size     = "Tracts in bin"
  ) +
  theme_minimal()
```

> **CHECKPOINT.** Dots should drift gently downward as the ratio rises, with a
> visible *step down* right at the dashed line. That step is the 8-percentage-point
> jump we planted.

## 2.3 Measure the jump with a regression

The simplest honest version: keep tracts *near* the cutoff (a ±10-point window),
and let the outcome depend on (a) a dummy for being below the cutoff and (b) a
straight line in the running variable that's allowed to differ on each side. The
dummy's coefficient is the jump.

```{r rd-reg-fake}
window <- fake |>
  filter(mfi_ratio >= 0.70, mfi_ratio <= 0.90) |>
  mutate(dist = mfi_ratio - 0.80)         # center the running variable at the cutoff

first_stage_fake <- lm_robust(
  got_nmtc ~ below_cutoff + dist + below_cutoff:dist,
  data = window
)

tidy(first_stage_fake) |> select(term, estimate, std.error, p.value)
```

> **CHECKPOINT.** The `below_cutoffTRUE` estimate should be roughly **0.08**
> (within ~2 standard errors of it). If it is: your pipeline works, and you just
> ran your first regression discontinuity. 🎉

## 2.4 Two experiments to run before moving on

Don't skip these — each one teaches a thing you'll need to *argue about* later:

1. **Shrink the window** to ±5 points (`0.75`–`0.85`) and re-run. The estimate
   should stay near 0.08 but the standard error should grow. *Why?* (Fewer tracts.)
   This is the bandwidth trade-off every RD paper wrestles with.
2. **Placebo cutoff:** re-run the regression pretending the cutoff is 0.90
   (re-center `dist` there, window 0.85–0.95). The "jump" should be ≈ 0 and
   insignificant. *Why is this comforting?*

```{r experiments, eval=FALSE}
# Your code for experiments 1 and 2 here. Copy 2.3 and modify it.
```

---

# Part 3 — The real thing

Now the same pipeline on real data. Create a `data/raw/` folder next to this
notebook and put the files there as you get them. Every chunk below checks
whether its input exists, so the notebook still knits while files are missing.

## 3.1 Data piece #1: NMTC transactions

**Preferred:** Ian's slice of the CDFI Fund transaction file, from the pod's
`Data/` folder in Drive — `nmtc_slice.csv`, with columns
`tract_geoid, cde, year, qlici_amount, total_project_cost` (see its one-page data
dictionary). **Fallback:** the full public release from the
[CDFI Fund data page](https://www.cdfifund.gov/documents/data-releases) — an Excel
file; read it with `read_excel()` and `clean_names()`, then rename to the columns
above.

```{r load-nmtc}
nmtc_path <- "data/raw/nmtc_slice.csv"

if (file.exists(nmtc_path)) {
  nmtc_raw <- read_csv(
    nmtc_path,
    col_types = cols(tract_geoid = col_character())  # NEVER read GEOIDs as numbers!
  )
  glimpse(nmtc_raw)
} else {
  message("Put nmtc_slice.csv in data/raw/ — see README. Skipping for now.")
}
```

> **Why `col_character()` matters (first real-data trap):** a tract GEOID is an
> 11-digit *label* — 2 state + 3 county + 6 tract digits. Read as a number,
> `01001020100` (Autauga County, Alabama) silently becomes `1001020100` and will
> never match anything again. Rule: **IDs are text, not numbers.**

### Clean it

Real administrative data is messy on purpose-revealing dimensions. Work through
each of these and *write down what you find* (worksheet: "data" section):

```{r clean-nmtc}
if (exists("nmtc_raw")) {
  nmtc <- nmtc_raw |>
    clean_names() |>
    mutate(
      tract_geoid = str_pad(str_trim(tract_geoid), width = 11,
                            side = "left", pad = "0")
    )

  # Q1: How many rows? How many DISTINCT tracts? (Many deals share a tract.)
  n_rows   <- nrow(nmtc)
  n_tracts <- n_distinct(nmtc$tract_geoid)
  c(rows = n_rows, distinct_tracts = n_tracts)

  # Q2: What years are covered? Which years does FREEDMAN's sample cover?
  #     (Check Section 3 of the paper — his window is the program's early
  #      years. Filter to match HIS window and note the choice.)
  count(nmtc, year)

  # Q3: Any rows with missing or zero qlici_amount? What will you do with them?
  summary(nmtc$qlici_amount)
}
```

```{r filter-years}
if (exists("nmtc")) {
  paper_years <- c(2003, 2004, 2005, 2006, 2007)  # <-- CHECK against Section 3 and fix!
  nmtc_sample <- nmtc |> filter(year %in% paper_years)
}
```

> **CHECKPOINT.** After cleaning, every `tract_geoid` should be exactly 11
> characters: `all(nchar(nmtc_sample$tract_geoid) == 11)` must be `TRUE`.

## 3.2 Data piece #2: the running variable

The transaction file only contains tracts that *got* deals. For an RD you need
**all tracts** — including the thousands that got nothing — each with its income
ratio. Get the CDFI Fund's census-2000 NMTC eligibility file (all ~65k tracts with
MFI ratio, poverty rate, and eligibility flags); Novoco's
[NMTC resource center](https://www.novoco.com/resource-centers/new-markets-tax-credits)
hosts the same data if the CDFI page is hard to navigate. Save it as
`data/raw/eligibility_2000.csv` (export the Excel sheet to CSV if needed) with at
least: `tract_geoid`, `mfi_ratio` (as a proportion or percent — check!),
`poverty_rate`, and a metro/non-metro indicator if present.

```{r load-eligibility}
elig_path <- "data/raw/eligibility_2000.csv"

if (file.exists(elig_path)) {
  elig <- read_csv(elig_path, col_types = cols(tract_geoid = col_character())) |>
    clean_names() |>
    mutate(tract_geoid = str_pad(str_trim(tract_geoid), 11, "left", "0"))

  # Sanity checks — run and READ the output:
  nrow(elig)                       # ~65,000 (all census-2000 tracts)?
  summary(elig$mfi_ratio)          # centered near 1? Or near 100 (= percent)? Rescale if so!
  mean(elig$poverty_rate >= 0.20, na.rm = TRUE)  # plausible share of high-poverty tracts?
}
```

> **The OR-clause trap (the classic mistake in this replication).** Eligibility is
> *income ≤ 80% of area median* **OR** *poverty ≥ 20%*. A tract at 85% of area MFI
> with 22% poverty is **eligible** even though it's above the income cutoff. If you
> ignore this, your "ineligible" side is contaminated and the jump shrinks.
> **Go find how Freedman handles this** (hint: look at how he defines his sample
> around the income threshold) and do the same. Write down what he does — this is
> a worksheet "sample" entry, and it's the single most important sample decision
> in the project.

```{r or-clause}
if (exists("elig")) {
  # Most likely answer (verify against the paper!): keep tracts that are NOT
  # eligible via the poverty route, so the income cutoff is the only margin.
  elig_clean <- elig |> filter(poverty_rate < 0.20)
  nrow(elig_clean)
}
```

## 3.3 Merge: build the analysis file

One row per tract: its running variable + its NMTC dollars (zero if none). The
direction of the join is the whole game — the *eligibility universe* is the left
table, so untreated tracts survive with `NA`, which we then make an explicit 0.

```{r merge}
if (exists("elig_clean") && exists("nmtc_sample")) {
  tract_invest <- nmtc_sample |>
    group_by(tract_geoid) |>
    summarise(
      n_deals      = n(),
      total_qlici  = sum(qlici_amount, na.rm = TRUE)
    )

  analysis <- elig_clean |>
    left_join(tract_invest, by = "tract_geoid") |>
    mutate(
      n_deals     = replace_na(n_deals, 0),
      total_qlici = replace_na(total_qlici, 0),
      got_nmtc    = as.integer(n_deals > 0)
    )

  # Diagnostics — interrogate the merge before trusting it:
  mean(analysis$got_nmtc)                            # share of tracts with any deal (small!)
  n_unmatched <- nrow(tract_invest) -
    sum(tract_invest$tract_geoid %in% elig_clean$tract_geoid)
  n_unmatched   # deals in tracts NOT in your eligibility universe — WHY?
}
```

> **CHECKPOINT.** Some transactions *won't* match (`n_unmatched > 0`) — possible
> reasons: you dropped poverty-eligible tracts in 3.2 (fine — those deals are
> outside the experiment), 2010-vintage GEOIDs in later transaction years (which
> is itself a reason to match Freedman's early sample window), territories.
> Account for every unmatched deal in one of those buckets before moving on.
> "I can explain every row I dropped" is what separates a replication from a vibe.

## 3.4 Run the first stage

Identical shape to 2.3 — that was the point of the rehearsal.

```{r first-stage}
if (exists("analysis")) {
  window_real <- analysis |>
    filter(mfi_ratio >= 0.70, mfi_ratio <= 0.90) |>
    mutate(
      below_cutoff = mfi_ratio <= 0.80,
      dist         = mfi_ratio - 0.80
    )

  fs_any <- lm_robust(got_nmtc ~ below_cutoff + dist + below_cutoff:dist,
                      data = window_real)
  fs_amt <- lm_robust(total_qlici ~ below_cutoff + dist + below_cutoff:dist,
                      data = window_real)

  tidy(fs_any) |> select(term, estimate, std.error, p.value)
  tidy(fs_amt) |> select(term, estimate, std.error, p.value)
}
```

And the picture — the real version of 2.2:

```{r rd-plot-real, fig.width=8, fig.height=5}
if (exists("window_real")) {
  window_real |>
    mutate(bin = cut_width(mfi_ratio, width = 0.01)) |>
    group_by(bin) |>
    summarise(bin_mid = mean(mfi_ratio), share = mean(got_nmtc), n = n()) |>
    ggplot(aes(bin_mid, share)) +
    geom_point(aes(size = n), alpha = 0.7) +
    geom_vline(xintercept = 0.80, linetype = "dashed", color = "red") +
    scale_y_continuous(labels = percent) +
    labs(
      title = "The real first stage: share of tracts with any NMTC deal",
      x = "Tract MFI / area MFI", y = "Share with any NMTC deal", size = "Tracts"
    ) +
    theme_minimal()
}
```

---

# Part 4 — Compare and reconcile (the actual deliverable)

## 4.1 The comparison table

Fill in the paper's numbers (from the first-stage table you identified in Part 0)
and yours:

```{r comparison}
comparison <- tribble(
  ~quantity,                          ~freedman_2012, ~ours,  ~same_sign, ~similar_size,
  "Jump in P(any NMTC deal) at 0.80",  NA_real_,       NA_real_, NA,        NA,
  "Jump in NMTC $ at 0.80",            NA_real_,       NA_real_, NA,        NA,
  "N tracts in window",                NA_real_,       NA_real_, NA,        NA
)
comparison
```

**Success is not getting the identical number.** You're using a (possibly) newer
data release, your own sample rules, and a simpler specification. Success is:
same sign, same order of magnitude, **and a written list of reasons for every
remaining gap**.

## 4.2 The reconciliation checklist

For each gap, work through the suspects in this order (cheapest first):

1. **Sample window** — same years as the paper? Same bandwidth around the cutoff?
2. **Sample rules** — did you handle the poverty OR-clause the way he did? Metro
   and non-metro both in? Tracts with missing income?
3. **Data vintage** — his CDFI file was an early release; yours has revisions.
4. **Specification** — he may use local linear with a chosen bandwidth, controls,
   or clustered standard errors. Match one ingredient at a time and watch what moves.
5. **Units** — dollars vs. log dollars vs. an indicator; percent vs. proportion.

Document every experiment in the worksheet. *This section is the skill.* The
regression takes one line; knowing why two careful people get different numbers
from "the same" data is what the lab is actually training you to do.

---

# Part 5 — Stretch goals (rungs 4–5 of the 5-step plan)

Only after Part 4 is written up:

- **(a) One outcome.** Pick one neighborhood outcome around the cutoff (tract
  poverty or median home values from the 2005–2009 ACS) and estimate the same
  RD on it. You've now touched the paper's reduced form.
- **(b) The rural split.** Estimate the first stage separately for metro and
  non-metro tracts. *Does the eligibility discontinuity mobilize less money in
  rural tracts?* — if yes, you've produced a fact that feeds directly into Ian's
  rural mobilization-gap paper. (This is "finding the hole" — rung 4.)
- **(c) Build the running variable yourself** from Census 2000 SF3 (median family
  income, table P077) via the `tidycensus` package, and reconstruct the area
  medians (state for non-metro, max of state/metro for metro tracts). Comparing
  your homemade ratio to the CDFI file's is a serious data exercise.
- **(d) The Portugal question.** One page: what is the COMPETE2020 analogue of
  this design? What's the running variable? What's the cutoff? What data would
  Katia need? (This is rung 5 — your first research proposal.)

When you get here, duplicate the Notion worksheet page and write it up. Then we
talk about Baum-Snow & Marion (2009) — same skills, new program — or you keep
going on (d). Your call.
